Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-17 HinesHetlandGPR.tex
did more shit

@@

\section{Estimating Transient Strain Rates}\label{sec:Method}
We denote transient crustal strain rates as $\dot{\bm{\varepsilon}}(z)$, where $z$ is the ordered pair $(x,t)$, $x$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by spatially and temporally differentiating estimates of transient displacements, $\bm{u}(z)$.  As we discuss in Section \ref{sec:NoiseModel}, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$ because $\dot{\bm{\varepsilon}}$ is insensitive to common mode noise in GNSS data.  We make a prior assumption that each component of $\bm{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(z) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(z,z')$ is a covariance function indicating how we expect $u_i(z)$ to covary with $u_i(z')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = X(x,x')T(t,t'),
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS observations at position $x_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_{ij} = &u(x_i,t_j) + v(x_i,t_j) + w_{ij} + a^{(1)}_i + a^{(2)}_it_j + \\
         &a^{(3)}_i\sin(2 \pi t_j) + a^{(4)}_i\cos(2 \pi t_j) + a^{(5)}_i\sin(4 \pi t_j) + a^{(6)}_i\cos(4 \pi t_j), 
\end{split}
\end{align}
where $a^{(1)}_{i}$ is an offset that is unique for each GNSS monument, $a^{(2)}_{i}$ is the steady rate of tectonic deformation at $x_i$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $w_{ij}$ to denote normally distributed, uncorrelated noise. Correlated noise which does not have a parametric representation is denoted by $v$.  For example, $v$ can consist of temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we will only assume that $v \sim \mathcal{N}(0,C_v)$. We consider the six coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the tectonic deformation, $a^{(2)}_{i}$, is spatially correlated and we could invoke a tectonic model to form a prior on $a^{(2)}_{i}$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $a^{(2)}_{i}$ for each station, and there is no need to incorporate a prior. Likewise, the seasonal coefficients may be spatially correlated \citep{Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We now consider the column vector of $n$ GNSS observations, $\bm{d}_*$. Let $\bm{y}$ be the set of $(x_i, t_j)$ pairs describing where and when each of the GNSS observations have been made. Let $\bm{a}$ be the vector of coefficients from eq. (\ref{eq:Data}) for each of the $m$ GNSS stations. We use $\bm{P}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\bm{y}$. We  also denote the vector of uncorrelated noise for each observation as $\bm{w}$, whose standard deviations are given by the formal data uncertainty, $\bm{\sigma}$. The observations can then be viewed as a realization of the random vector
\begin{equation}
\bm{d} = u(\bm{y}) + v(\bm{y}) + \bm{w} + \bm{P}\bm{a},
\end{equation}
which is distributed as $\mathcal{N}(\bm{0},\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T)$, where
\begin{equation}\label{eq:Cd}
\bm{\Sigma} = C_u(\bm{y},\bm{y}) + C_v(\bm{y},\bm{y}) + 
              \mathrm{diag}\left(\bm{\sigma}^2\right).  
\end{equation}
It should be understood that the notation $f(\bm{x})$ and $f(\bm{x},\bm{y})$ is shorthand for $[f(x_i)]_{x_i \in \bm{x}}$ and $[f(x_i,y_j)]_{(x_i,y_j) \in \bm{x} \times \bm{y}}$, respectively.

The prior for each component of transient displacements is then conditioned with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_v$ have already been chosen. If $\kappa$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(z) = C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(z,z') = C_u(z,z') - C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}C_u(\bm{y},z').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

The posterior transients displacements, $\hat{\bm{u}}$, are found by evaluating eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for each displacement component. We differentiate the Gaussian process $\hat{\bm{u}}$ (see \cite{Abrahamsen1997}) to form an estimate of transient strain rate, $\dot{\bm{\varepsilon}}$, which is itself a spatially and temporally continuous Gaussian process. The components of the transient strain rate tensor, $\dot{\varepsilon}_{ij}$, are given by
\begin{equation}\label{eq:StrainRate}
\dot{\varepsilon}_{ij}(z) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(z)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(z)}{\partial x_i}\right).
\end{equation}
which are distributed as $\mathcal{N}(\mu_{\dot{\varepsilon}_{ij}},C_{\dot{\varepsilon}_{ij}})$, where
\begin{equation}\label{eq:StrainMean}
\mu_{\dot{\varepsilon}_{ij}}(z) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(z)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(z)}{\partial x_i} \right)
\end{equation} 
and  
\begin{equation}\label{eq:StrainCov}
C_{\dot{\varepsilon}_{ij}}(z,z') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(z,z')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(z,z')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no temporally correlated noise (i.e., $v = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{y}) + \bm{P}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section \ref{sec:Method}. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries.

To begin the algorithm, we let $\Omega$ be the index set of non-outliers in $\bm{d}_*$ and initiate it with all $n$ indices. This algorithm is iterative, and for each iteration we calculate the residual vector
\begin{align}\label{eq:Residual}
\bm{r} &= \frac{\bm{d}_* - \mathrm{E}\left[(u(\bm{y}) + \bm{P}\bm{a})|\tilde{\bm{d}}_* \right]}{\bm{\sigma}} \\
       &= \frac{1}{\bm{\sigma}}\left(\bm{d}_*  - 
          \left[\begin{array}{cc}
                C_u(\bm{y},\tilde{\bm{y}}) & \bm{P} \\
                \end{array}\right]
          \left[\begin{array}{cc}
                C_u(\tilde{\bm{y}},\tilde{\bm{y}}) + \mathrm{diag}(\tilde{\bm{\sigma}}^2) & \tilde{\bm{P}} \\
                \tilde{\bm{P}}^T  & \bm{0} \\
                \end{array}\right]^{-1}
          \left[\begin{array}{c}
                \tilde{\bm{d}}_* \\
                \bm{0} \\
                \end{array}\right]\right),
\end{align}
where the tilde indicates that only elements corresponding to indices in $\Omega$ are retained (e.g., $\tilde{\bm{y}} = \{y_i\}_{i\in\Omega}$). We then update $\Omega$ to be
\begin{equation}\label{eq:Update}
\Omega = \{i : |r_i| < \eta \tilde{\bm{r}}_{rms}\}, \ \ \ r_i \in \bm{r}
\end{equation} 
where $\tilde{\bm{r}}_{rms}$ is the root-mean-square of $\tilde{\bm{r}}$ and $\eta$ is an outlier tolerance. We use $\eta=4$ in this study, which seems to accurately identify outliers without unnecessarily decimating the data. Iterations continue until the new $\Omega$ is equal to the previous $\Omega$.

@@

\subsection{Noise model}\label{sec:NoiseModel}
Before we determine the transient strain rates, we must establish a prior for the transient deformation, $u$, and the noise, $v$. In this section we discuss our choice for the noise covariance function $C_v$. There have been numerous studies on temporally correlated noise in GNSS data \citep[e.g.,][]{Zhang1997,Mao1999,Williams2004,Langbein2008}. In these studies, temporally correlated noise was described with some combination of Brownian motion (also referred to as a Weiner process or a random walk), a first-order Gauss-Markov (FOGM) process, and/or flicker noise. There is some physical justification for using Brownian motion as a noise model because it accurately describes the power spectrum of motion resulting from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,Wyatt1989,Langbein1997}. Here we describe the time dependence of $v$ as a FOGM process and consider $v$ to be spatially uncorrelated. A FOGM process is a solution to the stochastic differential equation
\begin{equation}\label{eq:FOGMdef}
\dot{v}(t) + \alpha v(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process degenerates to the commonly used Brownian motion noise model under the condition that $\alpha=0$ and $v(0) = 0$, which is evident from the definition of Brownian motion \citep[e.g.,][]{Papoulis1991}. Our noise model which satisfies eq. (\ref{eq:FOGMdef}) is a Gaussian process with zero mean and covariance function
\begin{equation}\label{eq:FOGM}
C_v\left((x,t),(x',t')\right) = \frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right) \delta(||x - x'||_2). 
\end{equation}

>>>>

\newcommand*{\du}[1]{\underline{\underline{#1}}}

@@

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rate, denoted as $\dot{\du{\varepsilon}}(\zeta)$. We use $\zeta$ to represent the ordered pair $(\vec{x},t)$, where $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\du{\varepsilon}}(\zeta)$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\du{\varepsilon}}(\zeta)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(\zeta)$.  As we discuss in Section \ref{sec:NoiseModel}, estimates of $\dot{\du{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\vec{u}(\zeta)$ because $\dot{\du{\varepsilon}}$ is insensitive to common mode noise in GNSS data.  We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(\zeta) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(\zeta,\zeta')$ is a covariance function indicating how we expect $u_i(\zeta)$ to covary with $u_i(\zeta')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We henceforth drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t'),
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS observations at position $\vec{x}_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_{ij} = &u(\vec{x}_i,t_j) + \eta(\vec{x}_i,t_j) + w_{ij} + a^{(1)}_i + a^{(2)}_it_j + \\
         &a^{(3)}_i\sin(2 \pi t_j) + a^{(4)}_i\cos(2 \pi t_j) + a^{(5)}_i\sin(4 \pi t_j) + a^{(6)}_i\cos(4 \pi t_j), 
\end{split}
\end{align}
where $a^{(1)}_{i}$ is an offset that is unique for each GNSS monument, $a^{(2)}_{i}$ is the steady rate of tectonic deformation at $\vec{x}_i$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $w_{ij}$ to denote normally distributed, uncorrelated noise. Correlated noise which does not have a parametric representation is denoted by $\eta$.  For example, $\eta$ can consist of temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we will only assume that $\eta \sim \mathcal{N}(0,C_\eta)$. We consider the six coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the tectonic deformation, $a^{(2)}_{i}$, is spatially correlated and we could invoke a tectonic model to form a prior on $a^{(2)}_{i}$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $a^{(2)}_{i}$ for each station, and there is no need to incorporate a prior. Likewise, the seasonal coefficients may be spatially correlated \citep{Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We now consider the column vector of $n$ GNSS observations, $\bm{d}_*$. Let $\bm{\xi}$ be the set of $(\vec{x}_i, t_j)$ pairs describing where and when each of the GNSS observations have been made. Let $\bm{a}$ be the vector of coefficients from eq. (\ref{eq:Data}) for each of the $m$ GNSS stations. We use $\bm{G}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\bm{\xi}$. We  also denote the vector of uncorrelated noise for each observation as $\bm{w}$, whose standard deviations are given by the formal data uncertainty, $\bm{\sigma}$. The observations can then be viewed as a realization of the random vector
\begin{equation}
\bm{d} = u(\bm{\xi}) + \eta(\bm{\xi}) + \bm{w} + \bm{G}\bm{a},
\end{equation}
which is distributed as $\mathcal{N}(\bm{0},\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T)$, where
\begin{equation}\label{eq:Cd}
\bm{\Sigma} = C_u(\bm{\xi},\bm{\xi}) + C_\eta(\bm{\xi},\bm{\xi}) + 
              \mathrm{diag}\left(\bm{\sigma}^2\right).  
\end{equation}
It should be understood that notation such as $u(\bm{\xi})$ and $C_u(\bm{\xi},\bm{\xi})$ is shorthand for $[u(\xi_i)]_{\xi_i \in \bm{\xi}}$ and $[C_u(\xi_i,\xi_j)]_{(\xi_i,\xi_j) \in \bm{\xi} \times \bm{\xi}}$, respectively.

The prior for each component of transient displacements is then conditioned with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_\eta$ have already been chosen. If $\kappa$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(\zeta) = C_u(\zeta,\bm{\xi})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(\zeta,\zeta') = C_u(\zeta,\zeta') - C_u(\zeta,\bm{\xi})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}C_u(\bm{\xi},\zeta').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(\zeta) = \left[\begin{array}{cc}
                         C_u(\zeta,\bm{\xi}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(\zeta,\zeta') = C_u(\zeta,\zeta') - 
                    \left[\begin{array}{cc}
                          C_u(\zeta,\bm{\xi}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{\xi},\zeta') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(\zeta) = \left[\begin{array}{cc}
                         C_u(\zeta,\bm{\xi}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(\zeta,\zeta') = C_u(\zeta,\zeta') - 
                    \left[\begin{array}{cc}
                          C_u(\zeta,\bm{\xi}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{\xi},\zeta') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

We use eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to find the posterior east and north components of transient displacements. The posterior transient displacements are differentiated temporally and spatially to form an estimate of $\dot{\du{\varepsilon}}$. The components of $\dot{\du{\varepsilon}}$ are given by 
\begin{equation}\label{eq:StrainRate}
\dot{\varepsilon}_{ij}(\zeta) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(\zeta)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(\zeta)}{\partial x_i}\right),
\end{equation}
which are Gaussian processes with mean functions
\begin{equation}\label{eq:StrainMean}
\mu_{\dot{\varepsilon}_{ij}}(\zeta) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(\zeta)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(\zeta)}{\partial x_i} \right)
\end{equation} 
and covariance functions
\begin{equation}\label{eq:StrainCov}
C_{\dot{\varepsilon}_{ij}}(\zeta,\zeta') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(\zeta,\zeta')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(\zeta,\zeta')}{\partial x_i \, \partial x'_i} \right).
\end{equation} 
In the above equations we use $\hat{u}_i$ to denote posterior transient displacements along direction $i$ and $x_i$ to represent the components of $\vec{x}$.

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no temporally correlated noise (i.e., $\eta(\zeta) = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{\xi}) + \bm{G}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section \ref{sec:Method}. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries.

To begin the algorithm, we let $\Omega$ be the index set of non-outliers in $\bm{d}_*$ and initiate it with all $n$ indices. This algorithm is iterative, and for each iteration we calculate the residual vector
\begin{align}\label{eq:Residual}
\bm{r} &= \frac{\bm{d}_* - \mathrm{E}\left[(u(\bm{\xi}) + \bm{G}\bm{a})|\tilde{\bm{d}}_* \right]}{\bm{\sigma}} \\
       &= \frac{1}{\bm{\sigma}}\left(\bm{d}_*  - 
          \left[\begin{array}{cc}
                C_u(\bm{\xi},\tilde{\bm{\xi}}) & \bm{G} \\
                \end{array}\right]
          \left[\begin{array}{cc}
                C_u(\tilde{\bm{\xi}},\tilde{\bm{\xi}}) + \mathrm{diag}(\tilde{\bm{\sigma}}^2) & \tilde{\bm{G}} \\
                \tilde{\bm{G}}^T  & \bm{0} \\
                \end{array}\right]^{-1}
          \left[\begin{array}{c}
                \tilde{\bm{d}}_* \\
                \bm{0} \\
                \end{array}\right]\right),
\end{align}
where the tilde indicates that only elements corresponding to indices in $\Omega$ are retained (e.g., $\tilde{\bm{\xi}} = \{y_i\}_{i\in\Omega}$). We then update $\Omega$ to be
\begin{equation}\label{eq:Update}
\Omega = \{i : |r_i| < \lambda \tilde{\bm{r}}_{rms}\}, \ \ \ r_i \in \bm{r}
\end{equation} 
where $\tilde{\bm{r}}_{rms}$ is the root-mean-square of $\tilde{\bm{r}}$ and $\lambda$ is an outlier tolerance. We use $\lambda=4$ in this study, which seems to accurately identify outliers without unnecessarily decimating the data. Iterations continue until the new $\Omega$ is equal to the previous $\Omega$.

@@

\subsection{Noise model}\label{sec:NoiseModel}
Before we determine the transient strain rates, we must establish a prior for the transient deformation, $u$, and the noise, $\eta$. In this section we discuss our choice for the noise covariance function $C_\eta$. There have been numerous studies on temporally correlated noise in GNSS data \citep[e.g.,][]{Zhang1997,Mao1999,Williams2004,Langbein2008}. In these studies, temporally correlated noise was described with some combination of Brownian motion (also referred to as a Weiner process or a random walk), a first-order Gauss-Markov (FOGM) process, and/or flicker noise. There is some physical justification for using Brownian motion as a noise model because it accurately describes the power spectrum of motion resulting from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,Wyatt1989,Langbein1997}. Here we describe the time dependence of $\eta$ as a FOGM process and consider $\eta$ to be spatially uncorrelated. A FOGM process is a solution to the stochastic differential equation
\begin{equation}\label{eq:FOGMdef}
\dot{\eta}(t) + \alpha \eta(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process degenerates to the commonly used Brownian motion noise model under the condition that $\alpha=0$ and $\eta(0) = 0$, which is evident from the definition of Brownian motion \citep[e.g.,][]{Papoulis1991}. Our noise model which satisfies eq. (\ref{eq:FOGMdef}) is a Gaussian process with zero mean and covariance function
\begin{equation}\label{eq:FOGM}
C_\eta\left((\vec{x},t),(\vec{x}',t')\right) = \frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right) \delta(||\vec{x} - \vec{x}'||_2). 
\end{equation}

We constrain the hyperparameters for $\eta$, $\alpha$ and $\beta$, with a set of 38 continuous GNSS stations in Washington and Oregon which are east of 121$^\circ$ W.  These stations are sufficiently far from the subduction zone that they are unlikely to contain transient signal.  We clean the data for these stations by removing jumps at times of equipment changes, and we remove outliers that have been detected with the algorithm described in Section \ref{sec:Outlier}. We then find $\alpha$ and $\beta$ for each station timeseries with the Restricted Maximum Likelihood (REML) method \cite[e.g.,][]{Harville1974,Cressie1992}. The REML method finds the hyperparameters, which we collectively refer to as $\bm{\theta}$, that maximize the likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\bm{\theta}) = \left(\frac{\left|\bm{G}^T\bm{G}\right|}
                           {(2\pi)^{n-6m} 
                           \left| \bm{\Sigma}(\bm{\theta}) \right| 
                           \left| \bm{G}^T\bm{\Sigma}\bm{G} \right|}\right)^{\frac{1}{2}} 
                           e^{-\tfrac{1}{2}\bm{d}_*^T\bm{K}(\bm{\theta})\bm{d}_*},
\end{equation}
where
\begin{equation}
\bm{K}(\bm{\theta}) = \bm{\Sigma}(\bm{\theta})^{-1} - 
                      \bm{\Sigma}(\bm{\theta})^{-1}\bm{G}
         \left(\bm{G}^T\bm{\Sigma}(\bm{\theta})^{-1}\bm{G}\right)^{-1}
         \bm{G}^T\bm{\Sigma}(\bm{\theta})^{-1}.
\end{equation}
For this section, $\bm{d}_*$ is used to denote the displacement time series for just one station. We are also assuming $C_u(\zeta,\zeta')=0$ when estimating the noise hyperparameters for this section. \citet{Harville1974} showed that choosing the hyperparameters which maximize eq. (\ref{eq:REML}) is equivalent to choosing the hyperparameters which maximize the probability of drawing $\bm{d}_*$ from $\bm{d}$.  We use the REML method over the maximum likelihood (ML) method \citep[e.g.,][]{Langbein1997} because the REML method properly accounts for the improper prior that we assigned to $\bm{a}$. The distribution of inferred $\alpha$ and $\beta$ are shown in Figure \ref{fig:NoiseParams}. The amplitude of FOGM noise, $\alpha$, for the east and north components is notable low and are clustered around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\beta$ tend to cluster around 0 yr$^{-1}$, suggesting that Brownian motion would have also been an appropriate noise model. The amplitude of FOGM noise for the vertical component is significantly larger with a median value of 13.5 mm/yr$^{0.5}$. The inferred values for $\beta$ are higher for the vertical component with a median value of 8.21 yr$^{-1}$. In Figure \ref{fig:NoiseSamples}, we use the median values of $\alpha$ and $\beta$ to generate two random samples of FOGM noise. The samples span five years and over these five years the east and north samples drift by about 1 mm. In the context of detecting SSEs, which produce several mm's of surface displacement on the timescale of weeks, the estimated FOGM noise for the east and north component is negligible. In contrast, the estimated FOGM noise for the vertical component is larger than the signal we would expect from SSEs. We suspect that the higher amplitude for the FOGM model in the vertical component is accommodating for deficiencies in our rather simple seasonal model. Based on this analysis, we henceforth ignore temporally correlated noise in the east and north component because of its low amplitude. We also do not estimate $u$ for vertical deformation because of the low signal to noise ratio.
